---
title: "Project Report"
author: "YUE OUYANG 920241282"
date: "3/13/2022"
output: html_document
---

# 1.Abstract

The COVID-19 has been a global concerning topic and it has many affected factors, such as different country and efficiency of vaccination. This report aims to analyze the differences in seven-day mortality rate between different countries and to compare the degree of change in mortality rate before and after vaccination. In particular, this study calculates a tendency score to explore if there is a causal relationship between the different countries and COVID-19 mortality rate. This report aims to analyze average mortality rate with region and vaccination, I find that mortality rates varied across regions but were highest in each region at a particular time and the average mortality rate decreased in most regions after vaccine implementation.

# 2.Background

The discovery of the COVID-19 in Wuhan, Hubei province, in December 2019 began a global defense campaign. Since the virus mutated and spread quickly, many countries took notice about vaccine development. The current COVID-19 pandemic is caused by a corona virus and all viruses known to cause disease in humans belong to alpha-Coronavirus and beta-Coronavirus.The response to COVID-19 varies from country to country, and the development of COVID-19 vaccine are different in different country. Thus, to study the extent and rate of new COVID-19 infection, this report select many variables, such as new cases, region and first vaccine date. Our research focus on mortality rate in COVID-19 instead of cumulative cases because of interpret ability.

# 3.Introduction 

Since the outbreak of the COVID-19, people around the world have been actively preventing and controlling the COVID, and countries are developing vaccines. Vaccine efficiency has become an important factor in the fight against the COVID-19, because only by improving vaccine efficiency can we get to the root of this problem. The first question below is to more accurate analyze the regional impact on the COVID-19 and the mortality rate of different regions to speculate which region has the greatest impact, because the infection of the COVID-19 is different in different regions.

The second question below is to  analyzes the efficiency of vaccines by studying the difference in mortality rates before and after vaccination, which provides a better understanding of how vaccines work and a strong basis for encouraging universal vaccination.

1.Is there a relationship between region and COVID-19 mortality rate within global?

2.Does vaccine of COVID-19 reduce the number of mortality rate and how to influence, comparing the mortality rate before and after the COVID-19 vaccination? 

For the first question, we assume that the region will have a positive relationship with COVID-19 mortality rate and we hypothesize that vaccine has negative relationship with mortality rate.

# 4.Data Summary

We will explore the WHO COVID-19 data in this project. The data set from WHO Coronavirus and we will look at daily case number from 1/17/2020 to 3/2/2022 and contains 200 countries. The data set have many variables, such as county, WHO-region(WHO regional offices: WHO Member States are grouped into six WHO regions -- Regional Office for Africa (AFRO), Regional Office for the Americas (AMRO), Regional Office for South-East Asia (SEARO), Regional Office for Europe (EURO), Regional Office for the Eastern Mediterranean (EMRO), and Regional Office for the Western Pacific (WPRO)), new cases(New confirmed cases. Calculated by subtracting previous cumulative case count from current cumulative cases count) new deaths (New confirmed death cases. Calculated by subtracting previous cumulative death case count from current cumulative death cases count) and first vaccine data (Date of first vaccinations. Equivalent to start/launch date of the first vaccine administered in a country).

In this report,Since the graph shows that the number of new cases after vaccination is much more than the number before vaccination, which means that the vaccine has no effect. This conclusion is clearly wrong, thus response variable directly cannot be new cases. By dividing the number of new deaths by the number of new cases, I think that the resulting ratio, named mortality rate gives a better indication of the relationship with vaccine, in other words, it can better indicate whether the vaccine is working and how efficiency of vaccine. Then adding a new variable named avg_mortality_rate where each element represents the mean number of mortality rate for the previous 7 days (inclusive of the current day). As to why the seven-day classification is used, I believe that the COVID-19 has a seven-day incubation period, so this classification can be more accurate for analysis. Therefore, mortality rate is the response variable, and predictor variables are region and first vaccine data. Since we want to start the study with the day on which is confirmed diagnosis occurs, the data for new cases equal to 0 are removed and negative new cases and new deaths are removed as well.

```{r echo=FALSE, message=FALSE, warning=FALSE}
#data cleaning 
covid <- read.csv("https://covid19.who.int/WHO-COVID-19-global-data.csv")
vacc<-read.csv("/Users/1/Desktop/vaccination-data.csv",header = T)
colnames(covid)[1]="Date_reported"
library(dplyr)

#inner joint data
merged_data = vacc %>% filter(FIRST_VACCINE_DATE != "") %>% 
  right_join(covid, by = c("COUNTRY" = "Country")) %>%
  dplyr::select(COUNTRY, WHO_REGION, FIRST_VACCINE_DATE, Date_reported, New_cases,New_deaths) %>%
  mutate(started_vac = ifelse(FIRST_VACCINE_DATE > Date_reported, "no", "yes"))%>%
  filter(New_cases>=0 & New_deaths>=0)%>%
mutate(mortality_rate = ifelse(New_cases == 0, 0, New_deaths/New_cases*100)) 

#calculating the response variable:avg_mortality_rate
merged_data = merged_data %>% 
  group_by(COUNTRY) %>% 
  mutate(avg_mortality_rate = stats::filter(mortality_rate, filter = rep(1/7,7), side = 1)) %>%   na.omit()%>%
  filter(New_cases!=0)

range(merged_data$Date_reported)
length(unique(merged_data$COUNTRY))
```

```{r message=FALSE, warning=FALSE, include=FALSE}
library(ggplot2)
library(ggpubr)
library(gplots)
library(multcomp)
library(car)
library(gtsummary)
```

# 5.Descriptive Analysis
## 5.1 Daily cases analysis
The plot shows the trend of daily new cases in different regions, which indicates the difference of new cases in different regions. Also AMRO (Regional Office for the Americas) region was the first to have a large outbreak and still has high new cases so far, and there are two more concentrated outbreak times in this region, one around January 2021 and the other around the beginning of 2022. There are many reason for such results, for instance, the government does not pay attention to the prevention and control of COVID-19 and the prevalence of the COVID-19 vaccination is not high. For the WPRO ( Regional Office for the Western Pacific), it has been the lowest number of new cases among the six regions, showing a low infection rate even during the outbreak period. China and Russia are the two largest countries in the WPRO region,and Russia has a low rate of infections due to its climate, while China has effectively controlled the COVID-19 due to effective governmental control and community-wide disease prevention actions.

```{r, fig.width = 6, fig.height = 3, echo=FALSE, message=FALSE, warning=FALSE}
#plot daily cases in each region
ggplot(merged_data,aes(x=as.Date(Date_reported),y=New_cases))+geom_line(aes(color=WHO_REGION))
```

The plot shows the trend of daily new deaths in different regions, and it is very different from the previous new cases plot. Except for the WPRO region, the new deaths in the other fiver regions fluctuated greatly, especially in the AMRO and SEARO (South-East Asia) regions. And SEARO region probably produced the largest new deaths on record from April 2021 to July 2021 according to the plot above. The high level of illegal immigration in South-Easy Asia, which exacerbates the spread of the COVID-19, coupled with limited economic development and medical care that cannot keep up with the rate of infection, has led to a dramatic increase in new deaths over this period.

```{r, fig.width = 6, fig.height = 3, echo=FALSE, message=FALSE, warning=FALSE}
#plot daily deaths in each region
ggplot(merged_data,aes(x=as.Date(Date_reported),y=New_deaths))+geom_line(aes(color=WHO_REGION))
```

According to the plot about region and average mortality rate, I find that the mortality rate dropped significantly in most areas where the vaccine was administered, with the most significant drop in the EURO (Europe) region. However, mortality rates increased after vaccine implementation in AFRO (Africa), probably because of the lack of local economic development and low vaccine prevalence.

## 5.2 Average mortality rate analysis
```{r, fig.width = 6, fig.height = 3, echo=FALSE, message=FALSE, warning=FALSE}
#get the average of mortality rate in each region
region_rate<-merged_data%>%
  group_by(WHO_REGION,started_vac)%>%
  summarise(avg_mortality_rate=mean(mortality_rate))

#create a box plot of by average mortality rate and WHO_REGION
ggboxplot(region_rate,x="WHO_REGION",y="avg_mortality_rate")
```

From the box plot, we can see that the two regions which are SEARO and WPRO that have the lowest average mortality rate and EMRO(Europe) tends to has highest average mortality rate. The lower average mortality rate in these two regions is mainly due to the availability of sanitation and high vaccination rates in their own regions, while the EMRO region shows the highest average mortality rate porbably because of the population base and poor awareness of vaccination.

```{r,fig.width = 6, fig.height = 3, echo=FALSE, message=FALSE, warning=FALSE}
#histogram plot
ggplot(region_rate,aes(x=WHO_REGION,y=avg_mortality_rate))+
  geom_bar(aes(color=started_vac,fill=started_vac),stat="identity",position=position_dodge(width = 0.3))+
  xlab("REGION")+ylab("average mortality rate")
  
```

According to more visually study the difference in mortality rates in different regions, a plot like the below was made. It indicates that the period from January 2021 to July 2021 is the period of high mortality, with different regions producing massive mortality in this period to different degrees.

```{r, fig.width = 6, fig.height = 3, echo=FALSE, message=FALSE, warning=FALSE}
#spaghetti plot
merged_data%>%
  mutate(Date_reported = as.Date(Date_reported)) %>%
  group_by(WHO_REGION,Date_reported)%>%
  summarize(m_avg_mortality_rate=mean(avg_mortality_rate))%>%
  ggplot(aes(x=Date_reported,y=m_avg_mortality_rate,by=WHO_REGION))+
  geom_line(aes(color=WHO_REGION))+
  xlab("Date")+ylab("mean of mortality rate")
```

## 5.3 Specifically analyzing AMRO region in new cases and new deaths.

This plot shows that the United stated is the region with the highest number of new cases and new deaths. Most of the other regions had high new cases but lower mortality rates, probably because other regions had better protection and higher vaccine coverage.

```{r, fig.width = 6, fig.height = 3, echo=FALSE, message=FALSE, warning=FALSE}
#scatter plot of AMRO region
merged_data%>% 
  filter(Date_reported=="2021-01-28", WHO_REGION=="AMRO") %>% 
  ggplot(aes(x=New_cases,y=New_deaths)) +
  geom_point()+
  geom_text(aes(label=COUNTRY),hjust=0, vjust=0)

```

# 6.Inferential Analysis

## 6.1 Build a two-way ANOVA model

The model by using two-way ANOVA is $$Y_{ijk} = \mu_{..} + \alpha_{i} +(\alpha\beta)_{ij}+\epsilon_{ijk}$$
where the WHO_REGION (denoted factor A) has $a=6$ levels: AFRO($i=1$), AMRO($i=2$), SEARO($i=3$), EURO($i=4$), EMRO($i=5$), and WPRO($i=6$), and the started_vac (factor B) has $b=2$ levels :$j=1 \  and\  2$.$\mu_{..}$ is represented the over mean and $\mu_{..}=\sum_{i = 1}^{a}\sum_{j=1}^{b}\mu_{ij}/(ab)$. A total of 978417 subjects suffering from STAR are selected and each of the 9 compounds are given to randomly selected n=108713 individuals. It is natural to consider a two-way ANOVA model here, and the next question is whether the interaction effects are present. 

## 6.2 Assumptions about the model

The assumptions about two-way anova are below:

1.The response variable in each group basically follow normal distribution.

2.The variance of each group is essentially the same.

3.Observations are independent of each group and within-group observations are obtained from a random sample.

## 6.3 Transformation in response variable.
```{r, fig.width = 6, fig.height = 3, echo=FALSE, message=FALSE, warning=FALSE}
#Test response variable whether need transformation.
hist(merged_data$avg_mortality_rate,breaks = 100)
```

According to the histogram of response variable, I find that it need a transformation because of normality valid.

```{r, fig.width = 6, fig.height = 3, echo=FALSE, message=FALSE, warning=FALSE}
#Transformation 
log_y<-log(merged_data$avg_mortality_rate+0.02)
hist(log_y,breaks=100)
merged_data = merged_data %>% mutate(log_y=log(avg_mortality_rate+1/4))
```

## 6.4 Select model by using average mortailty rate 

Using F-test to solve the question below mentioned. Firstly. suppose that we are interested in testing the presence of interactions.
Null hypothesis is $$H_o:(\alpha\beta)_{ij}=0$$
Alternative hypothesis is $$H_1:not\ all\ (\alpha\beta)_{ij}\ are\ zero$$
Thus, $$\ Full\ model:Y_{ijk}=\mu_{..}+\alpha_i+\beta_j+(\alpha\beta)_{ij}+\epsilon_{ijk}$$
$$\ Reduced\ model:Y_{ijk}=\mu_{..}+\alpha_i+\beta_j+\epsilon_{ijk}$$
```{r echo=FALSE, message=FALSE, warning=FALSE}
#Exploratory analysis
merged_data$WHO_REGION=as.factor(merged_data$WHO_REGION)
merged_data$started_vac=as.factor(merged_data$started_vac)
options(repr.plot.width=12,repr.plot.height=12)
par(mfrow=c(2,2))
# main effect plot for WHO_REGION.
plotmeans(log_y~WHO_REGION,data=merged_data,xlab="WHO_REGION",ylab="Average mortality rate",main="Main effect, WHO_REGION",cex.lab=1.5)

#main effect plot for started_vac
plotmeans(log_y~started_vac,data=merged_data,xlab="started_vac",ylab="Average mortality rate",main="Main effect, started_vac",cex.lab=1.5)

#interaction plot
interaction.plot(merged_data$WHO_REGION,merged_data$started_vac,merged_data$log_y,cex.lab=1.5,ylab="average mortality rate",xlab="WHO_REGION")
par(mfrow=c(1,1))
```

```{r, fig.width = 6, fig.height = 3, echo=FALSE, message=FALSE, warning=FALSE}
#Test for interactions
full_model=lm(log_y~as.factor(WHO_REGION)+as.factor(started_vac)+as.factor(WHO_REGION)*as.factor(started_vac),data=merged_data)
reduced_model=lm(log_y~as.factor(WHO_REGION)+as.factor(started_vac),data=merged_data)
anova(reduced_model,full_model)
```

According to the Analysis of Variance Table, F=119.85 and p-value<2.2e^-16, we can reject the null hypothesis and interaction terms cannot be dropped.

```{r, fig.width = 6, fig.height = 3, echo=FALSE, message=FALSE, warning=FALSE}
#Fit the chosen model
anova.fit=aov(log_y~as.factor(WHO_REGION)+as.factor(started_vac)+as.factor(WHO_REGION)*as.factor(started_vac) ,data=merged_data)
summary(anova.fit)
tbl_regression(anova.fit,exponentiate=T)
 
```

## 6.5 Tukey mathod to test started_vac
```{r echo=FALSE, message=FALSE, warning=FALSE}
anova.fit2=aov(log_y~WHO_REGION+started_vac+WHO_REGION*started_vac,data=merged_data)
summary(glht(anova.fit2,linfct = mcp(started_vac ="Tukey")))
```
For the secondary question of interest, using Tukey-Kramer method to test the effect of vaccination.
The null hypothesis is $$\mu_1=\mu_2$$
The alternative hypothesis is $$ not\ all\ \mu \ are\ equal $$
From Simultaneous Tests for General Linear Hypotheses, I find that p-value<2e-16 is very small and it can reject the null hypothesis from linear hypothesis about yes-no==0 .Thus, it is concluded that the vaccine has some influence.

# 7.Sensitivity Analysis

## 7.1 diagnostic plots
```{r, echo=FALSE, message=FALSE, warning=FALSE}
par(mfrow=c(2,2))
plot(anova.fit,cex.lab=1.2)
```

The residual v.s. fitted values plot shows that the residuals and fitted values are indpendent. About the Q-Q plot, I find that the points fall roughly along the straight diagonal line, and some points deviate from the line at the tail ends, but not enough to declare that the residuals are non-normally distributed. The reason for this situation may be that the average mortality rate is a bit too large and small, and some more detailed analysis may needed in the future in order draw more accurate conclusions.

## 7.2 Test assumption 
```{r echo=FALSE, message=FALSE, warning=FALSE}
leveneTest(log_y~as.factor(WHO_REGION)*as.factor(started_vac) ,data=merged_data) 
```

Using levenetest to test whether the variance is equal in this model. The null hypothesis is $$ H0:\sigma_1^2= \sigma_2^2  $$
The alternative hypothesis is $$ Not\ all\ variances \ are\ equal. $$
According to the result table, The p-value <2.2e-16 and it indicates that we need to reject the null hypothesis and the variances are not all equal.

```{r,fig.width = 6, fig.height = 3, echo=FALSE, message=FALSE, warning=FALSE}
#define model residuals
resid<-anova.fit$residuals

#create histogram of residuals
hist(resid,main="Histogram of Residuals",xlab="Residuals")
```

The residuals are roughly not normally distributed, so we cannot assume the normality assumption is met.

To fit a reasonable data set, I delete all average mortality rate which is zero and the results are below:

```{r, echo=FALSE, message=FALSE, warning=FALSE}
#delete all zero in average mortality rate
new_data<-merged_data[merged_data$avg_mortality_rate>0,]
log_y2<-log(new_data$avg_mortality_rate+0.02)
hist(log_y2,breaks=100)

#Fit the new chosen model
anova.fit2=aov(log_y2~as.factor(WHO_REGION)+as.factor(started_vac)+as.factor(WHO_REGION)*as.factor(started_vac) ,data=new_data)
par(mfrow=c(2,2))
plot(anova.fit2,cex.lab=1.2)

#define model residuals
resid2<-anova.fit2$residuals
```
````{r}
#create histogram of residuals
hist(resid2,main="Histogram of Residuals",xlab="Residuals")

#test assumption
leveneTest(log_y2~as.factor(WHO_REGION)*as.factor(started_vac) ,data=new_data) 
```

From the histogram of response variable, I find that it is normal distribution and from the residual plot, it is also shows that the the residuals are roughly normally distributed, so we can assume the normality assumption is met. But the variance is not equal as well. Also, I think the data was reduced substantially and the original 110174 was reduced to 92758, which is completely undesirable.

# 8.Discussion

By analyzing the seven-day average mortality rate in relation to region, I find that mortality rates varied across regions but were highest in each region at a particular time. Also, some regions had multiple peak points, such as ARMO and SEARO. The high level of illegal immigration in South-Easy Asia, which exacerbates the spread of the COVID-19, coupled with limited economic development and medical care that cannot keep up with the rate of infection, has led to a dramatic increase in new deaths over this period.For the second question, I find that the average mortality rate decreased in most regions after vaccine implementation, except for the AFRO region. The mortality rates increased after vaccine implementation in AFRO (Africa), probably because of the lack of local economic development and low vaccine prevalence. 

In this report, some data were removed in the process of data cleaning, mainly for the sake of computational simplicity. But the accuracy of the analysis may not be guaranteed in this case, in other words, it will affect the results. And it can also be concluded from looking at the diagnostic plots that the residual plots are not very well, probably because the response variable is not normal distribution.

After conducting the test of assumption, I find that both normality and equal variance failed the test so that our results from the model are not very reliable. For the reasons why the hypothesis did not pass I think there are several points. Firstly, the defect of the data itself makes the regression unsatisfactory, and then, there is a bias when I choose the response variable. But by deleting the average mortality rate of 0, the data was reduced substantially, even though the hypothesis was passed, but there is not much practical significance, and the original 110174 was reduced to 92758, which is completely undesirable.

# 9.Appendix

GitHub link:https://github.com/7qiwx/207project/tree/main/Yue%20Ouyang
 
# Session info {-}

<span style='color:blue'>
Report information of your `R` session for reproducibility. 
</span> 


```{r}
sessionInfo()
```



