---
title: "Effect of vaccination on the rate of fatality over WHO regions"
author: "(Shuwen Hou, 920250024)"
date: "03/13/2022"
output:
  html_document: 
    number_sections: yes
---
```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.pos = 'H')
```


# Abstract
COVID-19 has affect our daily life and is slowing down the global economy, and there has been over 16 million deaths so far. In this case, the government of countries had issued some policies such as wear a mask in public, social distance and get a vaccine. The social distancing was realized by online learning at both the primary and college levels, remote working at home, non-contact food delivery, temporary closed dine-in restaurant. Since the government strongly recommend people to get a vaccine to alleviate the epidemic, the project will focus on the effect of vaccination on the rate of fatality and find is there one region that has the highest mean of fatality rate. By doing the statistical analysis, I concluded that the vaccination has a positive effect to the rate of fatality over six WHO regions and there exist one region has the highest mean of fatality rate.


# Introduction
In this project, the main question of interest is how the vaccination affect the the rate of fatality across six WHO regions. According to the CDC's Morbidity and Mortality Weekly Report between April 4- December 25, 2021, it stated,"During October–November, unvaccinated persons had 13.9 and 53.2 times the risks for infection and COVID-19–associated death, respectively, compared with fully vaccinated persons who received booster doses, and 4.0 and 12.7 times the risks compared with fully vaccinated persons without booster doses"(paras.1). Base on all information, the possible hypothesis in this project is with the applying of vaccines, the fatality rate over WHO regions has a decreasing trend, if so, the second question of interest is if there is one region where the mean average fatality rate is the highest? 

## Target population

Since this project is focus on how vaccination affect the rate of fatality, there are too many target populations can be chosen as another level variable. However, using countries as our another variable will cause huge amount of observations and that is not good for conducting plots and doing descriptive analysis. In this case, using WHO regions as another variable is a good idea since there are total of 6 WHO regions, and also each region contains many neighboring countries, and it does not greatly affect the accuracy of our results. 

## Data and Variables
All datasets in this project are all from World Health Organization, The datasets will be updated daily, which let the project more accurate and reasonable. I use the mean of every seven days' fatality rate as my response variable since deaths occurs might take few days after the virus were detected. The other main variables I will use are:

* WHO_Region: I have total of 6 WHO regions -- Regional Office for Africa (AFRO), Regional Office for the Americas (AMRO), Regional Office for South-East Asia (SEARO), Regional Office for Europe (EURO), Regional Office for the Eastern Mediterranean (EMRO), and Regional Office for the Western Pacific (WPRO).
      
* Date_Updated: date of last updated.
      
* First_Vaccination_date: Date of first vaccinations. Equivalent to start/launch date of the first vaccine administered in a country.

* Country:I have total of 228 countries.Country, territory, area.

* Cumulative_cases:Cumulative confirmed cases reported to WHO to date.

* Cumulative_deaths: Cumulative confirmed deaths reported to WHO to date.

* v_exist: The existence of vaccination. This is a categorical predictor. "no" indicates the vaccines are not yet in use; "yes" indicates the vaccines are in use at that time.

* Fatality_rate: defined as: $fatality\ rate = \frac{New\ deaths}{New\ cases}$


# Background 

In these few year, COVID-19 is the most difficult challenge for human, it has caused a lot of impact on our study, life and work. COVID-19 is a respiratory disease caused by SARS-CoV-2, a corona virus discovered in 2019, at Wuhan city, Hubei Province, China. The virus that causes COVID-19 can spread through respiratory droplets when an infected person coughs, sneezes, or talks. The rapid spread of the virus has caused a very serious epidemic  around the world, which indicates COVID-19 is threatening human's lives. In this case, government advocates social-distance to prevent the spread of virus; afterwards, COVID_19 vaccines were developed to promote people to have antibodies to the virus thereby reducing the infection and mortality rate. As Mario Coccia(2020) said in article ***Factors determining the diffusion of COVID-19 and suggested strategy to prevent future accelerated viral infectivity similar to COVID ***, complex interaction between air pollution, meteorological conditions and biological characteristics of viral infectivity, and vaccination, health level of individuals such as habits, immune system, age, sex, both of them are possible factors that affect the transmission rate. In this case, the government has issued some policies such as wearing a mask in public, keeping 6 feet social distance and getting a vaccine. The social distancing was realized by online learning at both the primary and college levels, remote working at home, non-contact food delivery, temporary closed dine-in restaurant. 
The government strongly recommend people to get a vaccine and how vaccines can help this epidemic is the main motivation of this project.

```{r echo=FALSE, message=FALSE, warning=FALSE}

vaccination_data <-read.csv("/Users/taotaow/Desktop/vaccination-data.csv",header = T)
global_data <- read.csv("https://covid19.who.int/WHO-COVID-19-global-data.csv",header = T)

library(dplyr)
colnames(global_data)[3] <- "COUNTRY"
merge_data<-  vaccination_data %>% right_join(global_data,by="COUNTRY")
merge_data<- na.omit(merge_data)
merge_data <-merge_data[merge_data$New_cases >= 0, ]
merge_data <-merge_data[merge_data$New_deaths>= 0, ]

merge_data$fatality_rate <- ifelse(merge_data$New_cases =="0",0, merge_data$New_deaths/merge_data$New_cases*100)
merge_data$v_exist<- ifelse(merge_data$Date_reported > merge_data$FIRST_VACCINE_DATE,"yes", "no")

merge_data$ave_fatality_rate <- stats::filter(merge_data$fatality_rate, filter = rep(1/7, 7), sides = 1)
merge_data<- na.omit(merge_data)
merge_data <-merge_data[merge_data$New_cases!= 0, ]
library(dplyr)
m_data = dplyr::select(merge_data,Date_reported,WHO_region,COUNTRY,New_cases,New_deaths,v_exist,ave_fatality_rate)
m_data<- na.omit(m_data)


```

# Descriptive analysis 

Since there are huge amount variables in the original dataset, so it is instrumental to filter out the main variables, which can make following analysis much easier. After filtering the variables, deleting the missing value is the next important step. A new dataset with no missing values and containing multiple main variables obtained. Base on the background information and my assumption, it is necessary to find the vaccination starting date for each country and mark whether vaccine exists on each of reported date. However, the spread of virus may be delayed in some regions since it takes some time to spread from one country to another, therefore, after finding average fatality rate,deleting new cases which are zero from the final dataset is required since these data will lead to an inaccurate mean for all WHO region. 

By doing summary statistic, the table shows the mean of fatality rate over six region  for each situation(vaccine exists or not). Through observing the mean table, most of regions has a lower mean of fatality rate after the vaccine had been applied, for example, the AMRO region, when vaccines do not exist, the mean of fatality rate is 3.245707, which is higher than the mean of fatality rate when vaccines applied, which is 2.610602. However, there is one special region --AFRO, the mean of fatality rate when vaccines had been applied is higher than before. Poverty, poor medical healthcare and sanitation greatly explain how this happens. In order to compare them, a bar plot was conducted. From the bar plot, the vaccination has a significant influence on EURO region since there is a obvious difference between the mean_size with vaccination and the mean_size without the vaccination.

Mean of fatality rate on WHO_region with vaccination v.s without vaccination
```{r echo=FALSE, message=FALSE, warning=FALSE}
library(dplyr)

df1 <- data.frame(m_data %>%
  group_by(WHO_region,v_exist) %>%
  summarize(mean_size = mean(ave_fatality_rate)))
df1

```



```{r, fig.width = 6, fig.height = 3, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(gridExtra)
library(scales)
library(lubridate)
library(ggplot2)
ggplot(df1,aes(x=WHO_region,y=mean_size,fill= v_exist))+geom_bar(stat='identity',position='dodge')


```




Through conducting the spaghetti plot of average new deaths daily over six regions from the end of 2019 to March 2022, it is obvious that AFRO and WPRO regions maintain a relatively flat trend of mean of new deaths daily  by comparing with other regions, and also, these regions have a small number new deaths daily when look in long run. On the contrast, SEARO region has a huge amount average daily new deaths during May 2021 - August 2021 by compare with other regions. Moreover, the scatter plot of new cases of AMRO region was conducted, from the scatter plot, the highest daily new deaths is around 9000, and also there are some dots have very high amount of new cases.  
```{r , fig.width = 6, fig.height = 3,echo=FALSE, message=FALSE, warning=FALSE}
fig.spaghetti.1 <- m_data %>% 
  group_by(WHO_region, Date_reported)%>%
  summarise(m_new_deaths = mean(New_deaths)) %>%
  ggplot(aes(x=as.Date(Date_reported),y= m_new_deaths,by=WHO_region)) +
  geom_line(aes(color=WHO_region)) 
fig.spaghetti.1 

```

After obtain the spaghetti plot, conducting a scatter plot to show the new cases v.s new deaths of each country in one region on a particular day is help to find the statistics data in a region. In this case, I want to know the data in EMRO region on date 2021-11-28, and the plot clearly states Philippines has the highest new deaths, which is around 180. However, Viet Nam has the largest number of new cases, which is around 15000, and it also has a high new deaths daily, which is around 150. By using this method, we find easily find the new cases and new deaths of countries in different regions on a particular date.

```{r , fig.width = 6, fig.height = 3,echo=FALSE, message=FALSE, warning=FALSE}
fig.scatter.2 <- m_data %>% 
  filter(Date_reported=="2021-11-28",WHO_region=="WPRO") %>% 
  ggplot(aes(x=New_cases,y=New_deaths)) +
  geom_point(aes(color = COUNTRY))+
  geom_text(aes(label=COUNTRY),hjust=0, vjust=0)
  
fig.scatter.2
```


Let's draw a histogram of the response variable ro see its distribution. In order to clearly see the plot, I set breaks equal to 200, it is obvious that the response variable is not normally distributed, it is heavy right skewed. To solve this problem, it will be much better to take a log transformation to the response variable, which also will let the Normal Q-Q plot looks much better.
```{r message=FALSE, warning=FALSE}
#histogram of avg death rate
par(mfrow = c(1,2))
hist(m_data$ave_fatality_rate,  breaks = 500, xlab = 'Case-fatality Rate', main = '')
m_data$tave_fatality_rate = log(m_data$ave_fatality_rate + 0.00001)
hist( log(m_data$tave_fatality_rate + 0.00001),  xlab = 'transformed Case-fatality Rate', main = '')
```


# Inferential analysis 
We can define a two-way ANOVA model as follows: $Y_{ijk}$=$\mu_{ij}$+$\alpha_{i}$+$\beta_{j}$+$\epsilon_{ijk}$, where the index i represents the class type: AFRO (i=1), AMRO (i=2), EMRO (i=3), EURO (i=4), SEARO (1 =5), WPRO (i=6) and the index j represents the existence of vaccination (yes: exist vaccination; no: otherwise. The $\mu_{ij}$ indicates  theoretical mean or expected value of every seven days' fatality rate in cell (i, j), and the $\epsilon_{ijk}$ are i.i.d. N(0,$\sigma^2$). There are $\alpha\beta$ unknown means and one unknown σ in the cell means model.Write down the factor effect form with constraints:$Y{ijk}$=$\mu_{..}$+$\alpha_{i}$+$\beta_{j}$+ $(\alpha\beta)_{ij}$+$\epsilon_{ijk}$, i = 1,2,3,4,5,6, and j = yes/no. 

To find the best model, it is essential to compare the full model and reduced model to decide whether interaction term has a significant effect to the model or not. The first step is to set a full model and a reduced model. And second, use anova function to compare the full model and reduced model. Through observing the ANOVA table,  I found that the interaction term is significant to this model since this term has a extremely small p-value, which is smaller than 0.05. Therefore, dropping the interaction term is not recommended, we should keep the interaction term.

```{r echo=FALSE, message=FALSE, warning=FALSE}
full_model <- lm(tave_fatality_rate ~as.factor(WHO_region) +  as.factor(v_exist) +   as.factor(v_exist) *  as.factor(WHO_region), data = m_data)
#summary(full_model)

reduced_model <-  lm(tave_fatality_rate ~as.factor(WHO_region)+ as.factor(v_exist) , data = m_data)
#summary(reduced_model)
anova(reduced_model, full_model)
#final model
anova.fit <- aov(tave_fatality_rate~ as.factor(WHO_region)+ as.factor(v_exist) +   as.factor(v_exist) *  as.factor(WHO_region), data = m_data)


```
## F_test
doing a F test will help me analyze this model in more depth. First, test the main effects on WHO region term. The $\alpha_{1}$, $\alpha_{2}$ ,$\alpha_{3}$,$\alpha_{4}$, $\alpha_{5}$, $\alpha_{6}$ separately as, for AFRO (i=1), AMRO (i=2), EMRO (i=3), EURO (i=4), SEARO (1 =5), WPRO (i=6). First, we should state the null hypothesis, which is $$H_{0}: \alpha_{i} = 0, for\ all\ i\ ~vs~ H_{1}: not\ all\ \alpha_{i}\ are\ equal\ ,i =1,2,3,4,5,6 $$. For $\alpha{i}$, under the null hypothesis, the F value is  875.93, and the corresponding p=value is $2.2* 10^{-16}$, which is smaller than 0.05, therefore, the test indicates the main effects on WHO region can not be dropped.

Second, do a test on vaccination term,  which is $$H_{0}: \beta_{j} = 0, for\ all\ j\ ~vs~ H_{1}: not\ all\ \beta_{j} =0$$. For $\alpha{j}$, under the null hypothesis, the F value is  426.71, and the corresponding p=value is $2.2* 10^{-16}$, which is smaller than 0.05, therefore, the test indicates the main effects on vaccination term(v_exist) can not be dropped.

```{r}
anova(full_model)
```




#Equal Variance Test

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Calculate the variances for each group:
idx=m_data$WHO_region;
(var.comb=tapply(m_data$ave_fatality_rate, INDEX=idx,var))

```


The result indicates we should reject the null hypothesis at significance level $\alpha$ = 0.05. The variance of fatality rate of AFRO region is 15.79933; 25.08592 for AMRO region, 48.11878 for EMRO region, 49.74581 for EURO region,20.79747 for SEARO, 35.11353 for WPRO region. By comparing these numbers, it is obvious that we should reject the null hypothesis, which means not all variance for each region are equal.

Moreover, doing a Levene test is another good idea to justify if samples have equal variances. Define the null hypothesis  $H_{0}$: $\sigma_{1}^2$, $\sigma_{2}^2$ ,$\sigma_{3}^2$,$\sigma_{4}^2$, $\sigma_{5}^2$, $\sigma_{6}^2$  against the alternative  $H_{1}$: not all variance are equal. After conducting the Levene test, the p value is very small, less that 0.05, which indicates we should reject the null hypothesis at significance level $\alpha$ = 0.05. Therefore, we conclude that not all variance are equal.

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(car)
leveneTest(ave_fatality_rate~ v_exist*WHO_region, data = m_data)
```


## Tukey's Range Test

It seems that both ingredients and their interaction may have (statistically) significant impact on the response variable. We may want to explore what the best combination of ingredients is, or whether such combination exists. The search of "best combination" can be translated as finding the cell with the highest cell mean compared to other cells, which leads us to the Tukey-Kramer method. By doing the Tukey test at alpha equal to 5%, the highest mean of average fatality rate is in EMRO region, and the second largest is in AMRO region. 

```{r echo=FALSE, message=FALSE, warning=FALSE}
idx=m_data$WHO_region;
(means.comb=tapply(m_data$ave_fatality_rate, INDEX=idx,mean))
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
alpha=0.05;
#df1 <- data.frame(m_data$WHO_region)

T.ci=TukeyHSD(anova.fit,conf.level = 1-alpha)
T.ci$`as.factor(WHO_region)`
```


# Sensitivity analysis 

After taking transformation to the response variable, and fit in the model the Normal Q-Q plot was conducted, in this case, the Normal Q-Q plot looks better than the original full model. Even though there still a lots data that violate the normality assumption, the majority middle points follow the normality assumption. Since there are huge amount of zero new cases and new deaths at the beginning of the epidemic, and this situation will greatly affect the distribution of our data. Moreover, the rest of dots follow the normality assumptiom, which indicates our full model is still credible. 

```{r , fig.width = 6, fig.height = 3,echo=FALSE, message=FALSE, warning=FALSE}
par(mfrow=c(1,2))
anova.fit_1 <- aov(log(ave_fatality_rate+1/4) ~ as.factor(v_exist) + as.factor(WHO_region) +   as.factor(v_exist) *  as.factor(WHO_region), data = m_data)
#summary(anova.fit_1)

plot(anova.fit_1,cex.lab=1.2,which=1:2)

```



# Discussion 

This project is designed to find the how vaccines affect the average fatality rate in each WHO region, which can be explained as is there any difference between the average fatality rate over six WHO regions when vaccines exist and when vaccines do not exist. Based on my analysis from various aspects, I found the most region has a lower average fatality rate with the vaccination. There is one special region -- AFRO, has a higher average fatality rate when vaccines had been applied. This abnormal case can e explained as the poverty, poor medical healthcare of AFRO region can not directly let people see the effect of the vaccination. Therefore, the main question can be answered as following: When vaccines had been applied, the average fatality shows a decreasing trend over WHO regions. Moreover, through conducting the Tukey's range test, I found there exist a WHO region, EMRO, has a highest mean of average fatality, this finding answered my second question. 

However, in the process of conducting the diagnostic plot, I found that the Normal Q-Q plot shows the most of dots are not normally distributed, which indicates the response variable, average fatality rate need to be transformed. After take log transformation to the response variable, even though most of the points are normality distributed, there still exist some points violates the normality assumption. Therefore, this case can be a point to be improved in the future. 

Through doing analysis in this report, I found the vaccination do make a big contribution to alleviate the spread of virus and reduce the fatality rate, so in the future, government should allocate more money on medical healthcare area such as the research on vaccines. 


# Acknowledgement {-}


# Reference {-}


Johnson AG, Amin AB, Ali AR, et al. COVID-19 Incidence and Death Rates Among Unvaccinated and Fully Vaccinated Adults with and Without Booster Doses During Periods of Delta and Omicron Variant Emergence — 25 U.S. Jurisdictions, April 4–December 25, 2021. MMWR Morb Mortal Wkly Rep 2022;71:132–138. DOI: http://dx.doi.org/10.15585/mmwr.mm7104e2

Mario Coccia,Factors determining the diffusion of COVID-19 and suggested strategy to prevent future accelerated viral infectivity similar to COVID, Science of The Total Environment, Volume 729, 2020, 


# Session info {-}

```{r}
sessionInfo()
```

# Appendix {-}

```{r ref.label=knitr::all_labels(), echo=TRUE, eval=FALSE}

```